---
title: "Collate_rfp_cds_counts[20230523]"
output: html_notebook
---

Load all the required packages:
```{r}
library(riboWaltz)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
```

This notebook processes .bam files generated using STAR alignment against primary human GRCh38 GENCODE annotation.
.bam files used are in ./data/Try2
Trying with subsampled data i ./data/Try3

I am using riboWaltz to process this data.
I create annotation_dt using the human primary assembly basic annotation that I used to align my RFP data with STAR. I'm also loading the data and convert them into list.

```{r}
annotation_dt=create_annotation(gtfpath = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf")
bam_list <- bamtolist("data/Try3", annotation = annotation_dt, refseq_sep = "|")
```

I don't filter the reads because doing so causes me to lose 99% of the reads..
The periodicity in this data is not good.

```{r}
# Try filter only read lengths that satisfy a periodicity threshold
# (i.e. 70% of the reads fall in one of the three reading frames along the CDS)
#filtered_bam_list <- length_filter(data = bam_list,
#                                   length_filter_mode = "periodicity",
#                                   periodicity_threshold = 70)
# This doesn't seem to work? It threw away 99.9% of the reads. The reads kept have much longer length.
# Apparently it is because I don't have good periodicity to begin with. RNAse or CHX might be an issue.
```

Call for psite, flanking = 6, extremity = "auto"
Use default recommended settings:
Also save the offsets per sample in a text file

```{r}
offsets <- psite(data = bam_list)
samples <- unique(offsets$sample)
for (i in 1:length(samples)) {
  sam_offs <- subset(offsets, sample == samples[i], 
                    select = c("length", "corrected_offset_from_3"))
  colnames(sam_offs) <- c("length", "p_offset")
  write.table(sam_offs, paste("output/Try2/", samples[i], "_p-site-offsets.tsv", sep=""), 
              sep = "\t", row.names = FALSE)
}
```

Get the main table of reads with psite information.
Tried saving each psite list for each sample, but the filesize is way too big. Don't do it.

```{r}
reads_psite_list <- psite_info(bam_list, offsets, 
                               fastapath = "data/Try2/GRCh38.primary_assembly.genome.fa", 
                               fasta_genome = TRUE, refseq_sep = " ",
                               gtfpath = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf")

# reads_psite_list <- psite_info(bam_list, offsets, site=c("psite", "asite", "esite"),
#                                fastapath = "data/Try2/GRCh38.primary_assembly.genome.fa",
#                                fasta_genome = TRUE, refseq_sep = " ",
#                                gtfpath = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf")
#for (i in 1:length(reads_psite_list)){
#  sample_name <- names(reads_psite_list[i])
#  write.table(reads_psite_list[i], paste("output/Try2/", sample_name, "_p-site-list.tsv", sep=""), 
#              sep = "\t", row.names = FALSE)
#}

```

Check how the reads are distributed for a given transcript.

```{r}
# To plot read counts against position given a particular gene in a given sample:
transcript_id = "ENST00000398594.4"
reads_psite_sample = "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out"

# Get the read counts for a given p-siste position of a specific transcript in a given sample.
psite_counts <- reads_psite_list[[reads_psite_sample]][
  reads_psite_list[[reads_psite_sample]]$transcript == transcript_id ,
  ]$psite

# Convert counts table to data frame, change the column names.
counts <- as.data.frame(table(psite_counts))
colnames(counts) <- c("Position", "ReadCount")

# Generate a sequence of numbers from 0 or min. number to the max. number.
counts$Position <- as.numeric(as.character(counts$Position))
number_range <- seq(pmin(min(counts$Position), 0), max(counts$Position))

# Complete the data frame with missing numbers and set their counts to 0.
# Normalize ReadCount with the size of the library.
nrow_reads = nrow(reads_psite_list[[reads_psite_sample]])
counts_complete <- counts %>%
  complete(Position = number_range) %>%
  replace_na(list(ReadCount = 0)) %>%
  mutate(ReadCount = ReadCount / nrow_reads * 1000000)

# Parse information about genotype, rfp/rna, and rep number.
genotype <- ""
if (grepl("wt", reads_psite_sample)) genotype = "wt" else genotype = "Nterm"
rfp_or_rna  <- ""
if (grepl("rfp", reads_psite_sample)) rfp_or_rna = "RFP" else rfp_or_rna = "RNA"
rep_number <- ""
if (grepl("rep", reads_psite_sample)) {
  # Extract the number immediately following "rep"
  rep_number <- str_extract(reads_psite_sample, "(?<=rep)\\d+")
  }
# Combine the genotype, rfp/rna, and rep number information into a new sample name.
sample_name = paste(genotype, rfp_or_rna, rep_number, sep = "_")

p <- ggplot(counts_complete, aes(x = Position, y = ReadCount)) +
  geom_line() +
  labs(x = "position", y = "read count", title = paste(transcript_id, " read counts in ", sample_name), sep = "") +
  theme_bw() + 
  theme(panel.grid.minor = element_blank()) +
  coord_cartesian(ylim = c(0,1)) +
  scale_x_continuous(n.breaks=20)

#Limit max position to 5000 for ZNF562
#Limit max position to 500 for LLPH
#Limit max position to 350 for SNURF
#Limit max position to 300 for SEM1

#ggsave(filename = paste("output/Try2/", transcript_id, "_", sample_name, ".png", sep =""), plot = p, width = 8, height = 4, dpi = 600)
```

Checking for codon usage

```{r}
fasta_path = "data/Try2/GRCh38.primary_assembly.genome.fa"
gtf_path = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf"
sample_list <- c("wt_dep_rep2_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out")

wt1_wt2 <- c("wt_dep_rep1_RFP_Aligned.toTranscriptome.out", "wt_dep_rep2_RFP_Aligned.toTranscriptome.out")
wt2_wt3 <- c("wt_dep_rep2_RFP_Aligned.toTranscriptome.out", "wt_dep_rep3_RFP_Aligned.toTranscriptome.out")
wt1_wt3 <- c("wt_dep_rep1_RFP_Aligned.toTranscriptome.out", "wt_dep_rep3_RFP_Aligned.toTranscriptome.out")
Nt1_Nt2 <- c("Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out")
Nt2_Nt3 <- c("Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out")
Nt1_Nt3 <- c("Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out")
wt1_Nterm1 <- c("wt_dep_rep1_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out")
wt2_Nterm2 <- c("wt_dep_rep2_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out")
wt3_Nterm3 <- c("wt_dep_rep3_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out")


wt12_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt1_wt2, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
wt23_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt2_wt3, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
wt13_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt1_wt3, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
Nt12_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = Nt1_Nt2, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
Nt23_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = Nt2_Nt3, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
Nt13_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = Nt1_Nt3, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)

wtNterm1_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt1_Nterm1, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
wtNterm2_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt2_Nterm2, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)
wtNterm3_psite <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt3_Nterm3, 
                                fastapath = fasta_path, gtfpath = gtf_path, fasta_genome = TRUE, 
                                refseq_sep = " ", label_scatter = TRUE, label_number = 5, 
                                site = "psite", label_aminoacid = TRUE)

wt <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = wt_wt, fastapath = "data/Try2/GRCh38.primary_assembly.genome.fa", gtfpath = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf", fasta_genome = TRUE, refseq_sep = " ", label_scatter = TRUE, label_number = 10)


# trying1 <- codon_usage_psite(reads_psite_list, annotation = annotation_dt, sample = c("Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out", "wt_dep_rep1_RFP_Aligned.toTranscriptome.out"), fastapath = "data/Try2/GRCh38.primary_assembly.genome.fa", gtfpath = "data/Try2/gencode.v43.primary_assembly.basic.annotation.gtf", fasta_genome = TRUE, refseq_sep = " ", label_scatter = TRUE, label_number = 10)


# codon_coverage_example <- codon_coverage(reads_psite_list, annotation_dt, psite = FALSE)
# cds_coverage_example <- cds_coverage(reads_psite_list, annotation_dt)

#example_cu_barplot <- codon_usage_psite(reads_psite_list, 
#                                        annotation_dt, sample = #"wt_dep_rep1_RFP_Aligned.toTranscriptome.out",
#                                        fastapath = "path/to/transcriptome/FASTA/file",
#                                        fasta_genome = FALSE,
#                                        frequency_normalization = FALSE) 
#example_cu_barplot[["plot"]]
```
Checking read length distribution.
It seemed that for wild type average, I have longer read length overall compared to Nterm average.
Nevertheless, the median for both is between 28-29.

```{r}
sample_list <- list("wt_avg" = c("wt_dep_rep1_RFP_Aligned.toTranscriptome.out", "wt_dep_rep2_RFP_Aligned.toTranscriptome.out", "wt_dep_rep3_RFP_Aligned.toTranscriptome.out"), 
                    "Nterm_avg" = c("Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out"))

length_dist_dodged <- rlength_distr(bam_list, 
                                    sample = sample_list, cl = 99,
                                    multisamples = "average",
                                    plot_style = "dodged",
                                    colour = c("dodgerblue", "gray70"))
length_dist_dodged[["plot"]] 

```
Checking for trinucleotide periodicity.
Generally the P-site signal is a bit low..

```{r}
example_frames_stratified <- frame_psite_length(reads_psite_list, sample = "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out",
                                                region = "all", cl = 90)
example_frames_stratified[["plot"]]
```
Check for metaprofile psite as a representative of trinucleotide periodicity across all start codons.

```{r}
sample_list <- list("wt_avg" = c("wt_dep_rep1_RFP_Aligned.toTranscriptome.out", "wt_dep_rep2_RFP_Aligned.toTranscriptome.out", "wt_dep_rep3_RFP_Aligned.toTranscriptome.out"), 
                    "Nterm_avg" = c("Nterm_dep_rep1_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep2_RFP_Aligned.toTranscriptome.out", "Nterm_dep_rep3_RFP_Aligned.toTranscriptome.out"))

example_metaprofile_overlaid <- metaprofile_psite(reads_psite_list, annotation_dt, sample = sample_list, 
                                                  multisamples = "average", plot_style = "overlaid",
                                                  utr5l = 20, cdsl = 40, utr3l = 20,
                                                  frequency = TRUE, plot_title = "transcript",
                                                  colour = c("aquamarine4", "gray70"))
example_metaprofile_overlaid[["plot"]]
```


